---
title: "Mandibular shape in the Meso-Neolithic transition: the Zvejnieki case study"
author: "Maria Ana Correia"
date: '`r format(Sys.time(), "%d/%m/%Y")`'
subtitle: "Analyses Scripts"
output:
  html_document:
    code_folding: hide
    code_link: true
    keep_md: true
    toc: true
    toc_float: true
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    citation_package: default
    includes:
      in_header: mypackages.tex
bibliography: references.bib
link-citations: true
csl: apa-single-spaced
---

This file documents the geometric morphometrics of Zvejnieki mandibles. 3D landmarks were collected using 3DSlicer and statistical analyses used geomorph and SlicerMorph/SllicerMorphR [@adams2025; @rolfe2021].

```{r setup, echo=FALSE,include=FALSE}
#this makes images save in folder in directory
knitr::opts_chunk$set(
  echo = TRUE, #shows code
  warning = FALSE, message = FALSE, #stops warning messages
  fig.path = "images/",
  dev = c("svg", "png", "tiff"), #saves figures as svg, tiff, and png in images folder
  dpi = 500, #publishing quality for combination art (Elsevier)
  tidy.opts=list(width.cutoff=60), # stops code from running off page
  tidy=TRUE
)

# function to load or install from CRAN or GitHub
# installs missing packages when sharing with others
load_or_install <- function(pkg, github = NULL) {
  # If package not available, install it
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (is.null(github)) {
      # Install from CRAN
      install.packages(pkg, dependencies = TRUE)
    } else {
      # Install from GitHub (needs remotes package)
      if (!requireNamespace("remotes", quietly = TRUE)) {
        install.packages("remotes")
      }
      remotes::install_github(github, dependencies = TRUE)
    }
  }
  # Finally load it
  suppressPackageStartupMessages(
    library(pkg, character.only = TRUE)
  )
}

```

```{r packages, echo=FALSE,include=FALSE}
# CRAN packages
cran_pkgs <- c(
  "tidyverse", #everyday data analyses
  "styler", #source code formatter
  "arrow", #cross-language development platform to export .parquet
  "usethis", #automates repetitive tasks that arise during project setup
   "osfr", #interface for OSF
  "geomorph" #geometric morphometrics
  )

purrr::walk(cran_pkgs, load_or_install)

# GitHub packages (supply the repo as github = "user/repo")
load_or_install("SlicerMorphR", github = "SlicerMorph/SlicerMorphR")
#import SlicerMorph dataset into R


```

```{r functions, eval=FALSE,include=FALSE}
#!setting decimals
fmt_decimals <- function(decimals = 0) {
  function(x) format(x, nsmall = decimals, scientific = FALSE)
}
#graphical settings for ggplot
my_theme <- theme(
  axis.text = element_text(size = 8, colour = "black"),
  # makes numbers smaller and black (consider final display)
  axis.ticks = element_line(linewidth = 0.5,
                            colour = "black"),
  # same for ticks
  axis.title = element_text(size = 10),
  # and for axis titles
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_rect(
    colour = "black", fill = NA, size = 0.5)
)

#calculate outliers
is_outlier <- function(x) {
  return(
    x < quantile(x, 0.25, na.rm = TRUE) - 1.5 * IQR(x, na.rm = TRUE) |
      x > quantile(x, 0.75, na.rm = TRUE) + 1.5 * IQR(x, na.rm = TRUE)
  )
}

```

```{r variables, echo=FALSE, include=FALSE}

# ------------------------------------------------------
# CONFIGURATION
# ------------------------------------------------------
proj_node     <- osf_retrieve_node("vkat9")   # main project
raw_comp      <- osf_retrieve_node("kwafd")   # Raw Data component
derived_comp  <- osf_retrieve_node("9fnsp")   # Derived Data component

raw_dir <- "data/raw"            # raw JSONs folder
derived_dir <- "data/derived"    # derived outputs folder

usethis::use_directory(raw_dir)      # ensure folder exists
usethis::use_directory(derived_dir)

rds_path     <- file.path(derived_dir, "landmarks_array.rds")
csv_path     <- file.path(derived_dir, "landmarks_wide.csv")
parquet_path <- file.path(derived_dir, "landmarks_wide.parquet")

# ------------------------------------------------------
# DUAL-MODE LOGIC
# ------------------------------------------------------
if (nzchar(Sys.getenv("OSF_PAT"))) {
  
  # --- CASE A: PAT detected ---
  message("✅ PAT detected → processing raw JSON files from OSF")

  # Download JSONs into raw_dir
  json_files <- osf_retrieve_node(raw_comp) |>
  osf_ls_files() |>
  dplyr::filter(str_detect(name, "\\.json$"))
  osf_download(json_files, path = raw_dir, conflicts = "overwrite")
  
  # List all JSON files in raw_dir
json_files_paths <- list.files("data/raw", full.names = TRUE)

# Read JSONs and convert to numeric matrices, preserving rownames
landmark_list <- purrr::map(json_files_paths, function(f) {
  mat <- SlicerMorphR::read.markups.json(f)
  mat <- apply(mat, 2, as.numeric) # convert columns to numeric
  rownames(mat) <- rownames(SlicerMorphR::read.markups.json(f))
  mat
}) |> set_names(basename(json_files_paths))


# Build 3D array
p <- nrow(landmark_list[[1]])
k <- ncol(landmark_list[[1]])
N <- length(landmark_list)
array3d <- array(NA_real_, dim = c(p, k, N),
                 dimnames = list(rownames(landmark_list[[1]]),
                                 colnames(landmark_list[[1]]),
                                 names(landmark_list)))

# Fill the array slice by slice
for (i in seq_along(landmark_list)) {
  array3d[,,i] <- landmark_list[[i]]
}

  
  # Build 3D array: landmarks × XYZ × specimens
  p <- length(all_landmarks)
  k <- ncol(landmark_list[[1]])
  N <- length(landmark_list)
  array3d <- array(NA_real_, dim = c(p, k, N),
                   dimnames = list(all_landmarks, 
                                   colnames(landmark_list[[1]]), 
                                   names(landmark_list)))
  
  # Fill array slice by slice
  purrr::imap(landmark_list, ~ array3d[,,.y] <<- as.matrix(.x))
  
  # Create wide table for sharing
  df_wide <- purrr::imap_dfc(landmark_list, function(df, specimen) {
    df |> rowid_to_column("landmark") |>
      pivot_longer(X:Z, names_to = "axis", values_to = "value") |>
      unite("lm_axis", landmark, axis) |>
      pivot_wider(names_from = lm_axis, values_from = value) |>
      mutate(specimen = specimen)
  }) |> relocate(specimen)
  
  # Save outputs
  saveRDS(array3d, rds_path)
  write_csv(df_wide, csv_path)
  write_parquet(df_wide, parquet_path)
  
  # Optional upload to Derived Data
  tryCatch({
    osf_retrieve_node(derived_comp) |> osf_upload(path = derived_dir, conflicts = "overwrite")
  }, error = function(e){message("⚠️ Upload error: ", conditionMessage(e))})
  
} else {
  
  # --- CASE B: No PAT (reviewers) ---
  message("⚠️ No PAT → downloading derived CSV/Parquet from OSF")
  
  derived_files <- osf_retrieve_node(derived_comp) |>
    osf_ls_files() |>
    filter(name %in% c("landmarks_wide.csv","landmarks_wide.parquet")) |>
    osf_download(path = derived_dir, conflicts = "overwrite")
  
  message("✅ Downloaded: ", paste(basename(derived_files), collapse = ", "))
  
  # Load CSV or Parquet
  df_wide <- read_csv(csv_path)
  
  # Convert wide table back to 3D array using arrayspecs()
  specimen_ids <- df_wide$specimen
  landmark_cols <- names(df_wide)[-1]  # drop specimen column
  k <- 3
  p <- length(landmark_cols) / k
  n <- nrow(df_wide)
  
  array3d <- arrayspecs(as.matrix(df_wide[,-1]), p = p, k = k)
  dimnames(array3d)[[3]] <- specimen_ids
  
  message("ℹ️ 3D array reconstructed from derived CSV for reviewers")
}

```

```{r}

# ---- Dual-mode data retrieval ----
# This retrieves landmark data either:
# (A) from OSF raw JSON files (requires PAT), or
# (B) from derived dataset (.csv or .parquet) if no PAT is set.

if (nzchar(Sys.getenv("OSF_PAT"))) {
  message("🔑 Using OSF PAT: downloading raw JSON landmark files...")

  # 1. Retrieve and filter JSONs
  json_files <- osf_retrieve_node(raw_comp) |>
    osf_ls_files() |>
    dplyr::filter(str_detect(name, "\\.json$")) |>
    osf_download(path = "data/raw", conflicts = "overwrite")

  # 2. Parse JSONs into list of landmark matrices
  landmark_list <- map(json_files$local_path, SlicerMorphR::read.markups.json)

  # 3. Convert to 3D array (p × 3 × n)
  array3d <- arrayspecs(do.call(rbind, landmark_list),
                        p = nrow(landmark_list[[1]]),
                        k = 3)

} else {
  message("🟢 No OSF PAT detected: using derived dataset...")

  # 1. Retrieve derived files
  derived_file <- osf_retrieve_node(derived_comp) |>
    osf_ls_files() |>
    dplyr::filter(str_detect(name, "\\.(csv|parquet)$")) |>
    osf_download(path = "data/derived", conflicts = "overwrite")

  # 2. Load depending on extension
  if (str_detect(derived_file$name, "\\.csv$")) {
    df <- read.csv(derived_file$local_path)
  } else {
    library(arrow)
    df <- read_parquet(derived_file$local_path)
  }

  # 3. Convert to 3D array
  array3d <- arrayspecs(as.matrix(df),
                        p = length(unique(df$landmark)),
                        k = 3)
}

# Sanity check
print(dim(array3d))  # should be p × 3 × n

```

```{r tidy, echo=FALSE,eval=FALSE}

```

TODO: add an **OSF download step for reviewers**, so that when they run the Rmd they can skip the local GitHub file and pull the tidy `.RDS` directly from Derived Data

```{r stats, echo=FALSE}


```

# References

::: {#refs}
:::

```{r  session, echo=FALSE}
sessionInfo()
```
